---
title: "The effect of pH on wine volatile acidity"
author: "Claus Thorn Ekstrøm and Christian Bressen Pipper"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Volatile acidity (VA) refers to the process when lactic acid bacteria
and acetic acids turns wine into vinegar, and the process takes
place mainly due to growth of bacteria, the oxidation of ethanol, or
the metabolism of acids/sugars. Wines with a high level of pH are
supposedly more susceptible to oxidation and the antibacterial effects
of sulfur dioxide and of fumaric acid are reduced rapidly as the pH
level increases. Consequently, wines are thought to lose their quality
as they become less acidic (increased pH) since the volatile acidity
increases.

The paper by @cort:etal:2009 considers 11 physicochemical properties
of a selected sample of Portuguese *vinho verde* wines. Samples from
1599 red wines and 4898 white wines are available and the relationship
between volatile acidity and pH are shown separately for red and white
wines in the figure below.

```{r}
library(mommix)
library(magrittr)
data(wine)
## Remove extreme observations
wine2 <- wine %>% filter(residual.sugar<25)

## Plot the data
wine2 %>% ggplot(aes(x=pH, y=volatile.acidity, col=colour)) +
                geom_point(alpha = 1/5, position = position_jitter(h = 0), size = 2) +
		ylab("Volatile acidity") + 
                geom_smooth(method = 'lm', se=FALSE)
```

While it is apparent that the available red wines generally have
slightly higher levels of pH, it also appears as if the impact of pH
on volatile acidity is largest for the red wines in the sample:
Individual regression lines for the two types of wine show an almost
horizontal line for white wine while there is an effect of pH on VA
for the red wines:

```{r}
winelm2 <- lm(volatile.acidity ~ pH*colour-1, data=wine2)
summary(winelm2)
```


If we did not know that the full 6497 samples were comprised of two
different types of wine we might pursue regressing volatile acidity on
pH for the full data.

```{r}
winelm <- lm(volatile.acidity ~ pH, data=wine2)
summary(winelm)
```


This gives the dashed regression shown in the figure below, which
suggests an overall effect of pH on volatile acidity.


```{r}
wine2 %>% ggplot(aes(x=pH, y=volatile.acidity, col=colour)) + geom_point(alpha = 1/5, position = position_jitter(h = 0), size = 2) +
  ylab("Volatile acidity") + 
  geom_smooth(method = 'lm', se=FALSE) + 
  geom_abline(intercept=coef(winelm)[1], slope=coef(winelm)[2], linetype=2) 
```



When we fit the moment-based mixture model then we  see that the
data are likely to consist of a mixture, \(\hat{p} = 0.28\) (95\% CI:
0.24-0.33), which suggests that only a proportion of the wines are
influenced by pH:

```{r}
wineres <- moment_mixture_model(wine2, volatile.acidity ~ pH, weight="square")
wineres
```

(For now we use the following code, which only handles one predictor but provides the standard errors. This will be removed in the future)

```{r}
wineback <- cbpcode(wine2$volatile.acidity, wine2$pH)
wineback
```


Also, the effect driving the association between the
volatile acidity and pH appears to be much stronger than what was
observed from analyzing the full data with a simple linear regression
model, \(\hat\beta=\) 0.94 (95\% CI 0.51-1.37). Consequently, the moment
mixture model is able to identify that the wine data is likely to
consist of two types of wines that respond differently to changes in
levels of pH with only placing very few restrictions on the underlying
distributions.


